```{r}
#| echo: false
#| warning: false
#| message: false
#| collapse: true
library(reticulate)
if (Sys.getenv("GITHUB_ACTIONS") == "true") {
  # for running in GitHub Actions (otherwise reticulate can't find diverse_web_env)
  reticulate::use_python(Sys.getenv("RETICULATE_PYTHON"), required = TRUE)
} else {
  # for running locally
  reticulate::use_condaenv("diverse_web_env", required = TRUE)
}
```

### Key Features of the Dataset

The dataset offers rich, structured information about Indigenous-owned businesses in British Columbia. The key fields are:

-   **Business Name** (`business_name`) – The official name of the business. It serves as the primary identifier for each entry in the dataset.

-   **City** (`city`) – The city or town where the business is physically located.

-   **Latitude & Longitude** (`latitude`, `longitude`) – Geographic coordinates that allow for mapping and spatial analysis. Useful for geographic visualizations and regional studies.

-   **Region** (`region`)  – A broader classification of the business’s location within British Columbia (e.g., Northeast, Thompson/Okanagan). Helps in regional economic analysis.

-   **Type** (`type`) – Indicates the business ownership or structure, such as "Private Company," "Community owned" etc

-   **Industry Sector** (`industry_sector`) – The NAICS (North American Industry Classification System) code and description that categorize the business’s industry.

-   **Year Formed** (`year_formed`) – The year the business was established, which can be used to analyze trends in Indigenous entrepreneurship over time.

-   **Number of Employees** (`number_of_employees`)  – A range representing the workforce size, such as "1 to 4" or "5 to 9". This gives a sense of business scale and economic impact.

### Purpose and Use Case

The purpose of this document is to:

- Increase the visibility of Indigenous-owned businesses.
- Provide insights into the distribution, characteristics, and trends of Indigenous businesses in British Columbia.
- Supplemented by community data, study relationships between business presence and community characteristics.

## Case Study

### Objective

**Which regions in British Columbia have the highest density of Indigenous-owned businesses, and how do industry sectors vary across these clusters?**

The goal is to explore the geographic and economic distribution of Indigenous businesses using attributes such as region, city, and industry sector. We apply clustering techniques to identify natural groupings of businesses based on their geospatial coordinates and sectoral attributes. This helps highlight regional concentrations, suggest potential areas of economic activity, and offer insights that could inform future policy or investment considerations.

### Analysis

**Loading Libraries**

::: {.panel-tabset}
## R
```{r library_load, results = "hide", warning = FALSE, message = FALSE}
library(diversedata) 
library(tidyverse)  # Loads dplyr, tidyr, stringr, ggplot2, etc.
library(infer)      # For chi-square test
library(dbscan)     # For clustering
library(leaflet)    # For interactive maps
library(treemapify) # For creating treemaps
library(knitr)
```

## Python
```{python library_load_py, results = "hide", warning = FALSE, message = FALSE}
import diversedata as dd
import numpy as np
import pandas as pd
import altair as alt
import plotly.express as px
from scipy.stats import chi2_contingency
from sklearn.cluster import DBSCAN
from IPython.display import Markdown
np.random.seed(123) # set seed for reproducibility
```
:::

#### 1. Data Cleaning & Processing

To ensure the quality and consistency of the data set, several pre-processing steps were done:

-   **Removed records with missing geographic coordinates**: These records were excluded since latitude and longitude are crucial for spatial mapping and regional distribution analysis.

-   **Standardized the format of region names**: Region names were converted to title case to ensure consistent grouping and filtering.

-   **Converted field types**: The `Industry Sector` column was converted to a factor for categorical analysis, and `Year Formed` was converted to integer to allow for chronological trends and filtering.

::: {.panel-tabset}
## R
```{r data_load}
# Load and clean the data
indigenousbiz_data <- bcindigenousbiz |>
  filter(!is.na(latitude) & !is.na(longitude)) |>
  mutate(
    region = str_to_title(region),
    industry_sector = as_factor(industry_sector),
    year_formed = as.integer(year_formed)
  )

# Preview cleaned data
indigenousbiz_data |> 
  head() |>
  kable()
```

## Python
```{python data_load_py}
# Load and clean the data
indigenousbiz_data = dd.load_data("bcindigenousbiz")
indigenousbiz_data = indigenousbiz_data.dropna(subset=["latitude", "longitude"])
indigenousbiz_data["region"] = indigenousbiz_data["region"].str.title()
indigenousbiz_data["industry_sector"] = pd.Categorical(
    indigenousbiz_data["industry_sector"], ordered=False
)
indigenousbiz_data["year_formed"] = indigenousbiz_data["year_formed"].astype("Int64")

# Preview cleaned data
Markdown(indigenousbiz_data.head().to_markdown())
```
:::

#### 2. Exploratory Data Analysis (EDA)

**Business count by Region**

::: {.panel-tabset}
## R
```{r eda_1, warning=FALSE, message=FALSE}
# Business count by region with tidyverse style
indigenousbiz_data |>
  count(region, sort = TRUE) |>
  ggplot(aes(
    x = fct_reorder(region, -n),  # Use forcats for reordering
    y = n,
  )) +
  geom_col(fill = 'skyblue') +
  coord_flip() +
  labs(
    title = "Business Count per Region",
    x = "Region",
    y = "Number of Businesses"
  ) +
  theme_minimal()
```

## Python
```{python eda_1_py, warning=FALSE, message=FALSE}
alt.Chart(indigenousbiz_data).mark_bar().encode(
    x=alt.X("count()").title("Number of Businesses"),
    y=alt.Y("region").sort("x").title("Region"),
).properties(title="Business Count per Region", width=500, height=300)
```
:::

**Mapping Businesses**

::: {.panel-tabset}
## R
```{r eda_2, warning=FALSE, message=FALSE}
# Create leaflet map
indigenousbiz_data |>
  leaflet() |>
  addTiles() |>
  setView(lng = -125, lat = 54.5, zoom = 4.5) |>
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    label = ~business_name,
    radius = 3,
    color = "#1E90FF",
    stroke = FALSE,
    fillOpacity = 0.7
  )

```

## Python
```{python eda_2_py, warning=FALSE, message=FALSE}
#| output: false

fig_map = px.scatter_map(
    indigenousbiz_data,
    lat="latitude",
    lon="longitude",
    opacity=0.5,
    zoom=4,
    center={"lat": 54.5, "lon": -128},
    hover_name="business_name"
)

fig_map = fig_map.update_layout(
    margin=dict(l=0, r=0, t=30, b=10),
    width=700,
    height=550,
    mapbox_style="open-street-map"
)
```

```{python fig_map_html}
#| echo: false
html_fig_map = fig_map.to_html(include_plotlyjs='cdn', full_html=False)
```

```{r fig_map_show}
#| echo: false
library(htmltools)
HTML(py$html_fig_map)
```
:::

**Industry Analysis**

::: {.panel-tabset}
## R
```{r eda_3, warning=FALSE, message=FALSE}
# Plot top 10 industry sectors
indigenousbiz_data |>
  count(industry_sector, sort = TRUE) |>
  slice_max(n, n = 10) |>
  ggplot(aes(
    x = fct_reorder(industry_sector, -n),
    y = n,
  )) +
  geom_col(fill = 'skyblue') +
  coord_flip() +
  labs(
    title = "Top Industry Sectors",
    x = "Sector",
    y = "Number of Businesses"
  ) +
  theme_minimal()
```
## Python
```{python eda_3_py, warning=FALSE, message=FALSE}
top_10_sectors = (
    indigenousbiz_data.groupby("industry_sector", observed=True)
    .size()
    .reset_index(name="count")
    .sort_values(by="count", ascending=False)
    .head(10)
)

alt.Chart(top_10_sectors).mark_bar().encode(
    x=alt.X("count").title("Number of Businesses"),
    y=alt.Y("industry_sector").sort("x").title("Sector").axis(alt.Axis(labelLimit=300))
).properties(title="Top Industry Sectors", width=500, height=300)
```
:::

#### 3. Chi-Squared Test of Independence

To better understand the relationship between Indigenous business locations and their characteristics, we test whether `type` (ownership type) is independent of `region` using a Chi-square test.

**Hypotheses**

-   Null Hypothesis (H₀): Business ownership type and region are independent. That is, the distribution of ownership types is the same across all regions.

-   Alternative Hypothesis (Ha): Business ownership type and region are dependent. That is, the distribution of ownership types differs by region.

**Chi-squared Assumption Check**

-   Data in Frequency Form: The data used in this test are raw counts of business records, not percentages or proportions.

-   Categorical Variables: Both variables under study, type and region are categorical and represent discrete groups.

-   Independence of Observations: Each row in the data set corresponds to a unique business. There is no duplication, satisfying the assumption of independence.

-   Expected Count: The Chi-square test expects that all expected cell frequencies are ≥ 5. In our case this assumption is violated.

::: {.panel-tabset}
## R
```{r analysis_1, warning = FALSE}
#| results: hold

## Assumptions check

# Step 1: Chi-Square Assumption Check
type_region_table <- table(indigenousbiz_data$type, indigenousbiz_data$region)
chisq_result <- chisq.test(type_region_table)
expected_counts <- chisq_result$expected

min_expected <- min(expected_counts)

cat("Chi-Square Assumptions Check:\n")
cat("- Minimum expected count:", round(min_expected, 2), "\n")

if (min_expected >= 5) {
  cat("Chi-squared assumptions met. Proceeding with standard test.\n")
  print(chisq_result)
} else {
  cat("Assumptions violated. Proceeding with simulation-based Chi-squared test.\n\n")
}

```

## Python
```{python analysis_1_py}
#| results: hold

type_region_table = pd.crosstab(
    indigenousbiz_data["type"], indigenousbiz_data["region"]
)
chi2_stat, p_value, dof, expected = chi2_contingency(type_region_table)
min_expected = expected.min()

print("Chi-Square Assumptions Check:")
print(f"- Minimum expected count: {min_expected:.2f}")

if min_expected >= 5:
    print("Chi-squared assumptions met. Proceeding with standard test.")
    print(f"Chi2 Statistic: {chi2_stat}")
    print(f"P-value: {p_value}")
    print(f"Degrees of Freedom: {dof}")
else:
    print("Assumptions violated. Proceeding with simulation-based Chi-squared test.")
```
:::

**Simulation-based Chi-Square Test with `infer`**

Since the assumption regarding minimum expected cell counts was violated with at least one expected frequency falling below the threshold of 5. The Chi-squared test may yield inaccurate results. To overcome this, we use a simulation-based Chi-squared test via the infer package. This method generates a null distribution by permuting the data under the assumption of independence.

The simulation-based tests still require some assumptions to hold:

- The data must represent independent observations.

-   The variables must be categorical.

-   The permutations assume that under the null hypothesis, the distribution of values across categories is exchangeable.

::: {.panel-tabset}
## R
```{r analysis_2, warning = FALSE}
# Generate null distribution
null_distribution <- indigenousbiz_data |>
  specify(type ~ region) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "Chisq")

# Observed statistic
observed_stat <- indigenousbiz_data |>
  specify(type ~ region) |>
  hypothesize(null = "independence") |>
  calculate(stat = "Chisq")

# Plot permutation distribution
null_distribution |>
  visualize() +
  shade_p_value(obs_stat = observed_stat, direction = "greater")

# Compute p-value
p_val <- null_distribution |>
  get_p_value(obs_stat = observed_stat, direction = "greater")

cat("Simulation-based p-value:", round(p_val$p_value, 4), "\n")

```

## Python
```{python analysis_2a_py}
# Drop missing values in the columns of interest
df = indigenousbiz_data.dropna(subset=["type", "region"])

# Make function to calculate chi-square stat
def chi2_statistic(data):
    table = pd.crosstab(data["type"], data["region"])
    chi2, _, _, _ = chi2_contingency(table)
    return chi2

# Generate null distribution
reps = 1000
null_distribution = []

for rep in range(reps):
    shuffled = df.copy()
    shuffled["region"] = np.random.permutation(shuffled["region"])
    null_distribution.append(chi2_statistic(shuffled))

null_distribution = np.array(null_distribution)
null_df = pd.DataFrame({"chi2_stat": null_distribution})

# Observed statistic from earlier
observed_stat = chi2_stat

# Plot permutation distribution
histogram = alt.Chart(null_df).mark_bar(color="grey").encode(
    x=alt.X("chi2_stat:Q").bin(alt.Bin(maxbins=30)).title("Chi-square statistic"),
    y=alt.Y("count()").title("Count")
).properties(width=600, height=400, title="Simulation-based Null Distribution")

observed_stat_line = alt.Chart(pd.DataFrame({"obs": [observed_stat]})).mark_rule(color="red", size=3).encode(
    x="obs:Q"
)

permutation_dist = histogram + observed_stat_line
permutation_dist
```

```{python analysis_2b_py}
# Compute p-value (proportion of null stats >= observed stat)
p_value = np.mean(null_distribution >= observed_stat)

print(f"Simulation-based p-value: {p_value:.4e}")
```
:::

**Inference**

To assess whether there is an association between business type and region, we conducted a simulation-based chi-squared test using 1,000 permutations. The null distribution of test statistics generated under the assumption of independence shows that our observed statistic lies far in the tail of the distribution. This suggests that such an extreme result would be highly unlikely if business type and region were truly independent. Therefore, we reject the null hypothesis and conclude that business type and region are statistically dependent.

#### 4. Spatial Clustering of Indigenous Businesses using DBSCAN

To explore geographic patterns in Indigenous business, the technique of clustering based on latitude and longitude coordinates was applied. This analysis revealed several spatial clusters of Indigenous businesses across British Columbia. The treemap provided more information on which were the prominent industries in each cluster. 

::: {.panel-tabset}
## R
```{r analysis_3, warning = FALSE, message = FALSE}
## DBSCAN Clustering and Leaflet Map

coords <- indigenousbiz_data |>
  select(longitude, latitude) |>
  as.matrix()

set.seed(123)
db <- dbscan::dbscan(coords, eps = 1.2, minPts = 8)

indigenousbiz_data <- indigenousbiz_data |>
  mutate(cluster = factor(db$cluster))

pal <- colorFactor(palette = "Set1", domain = indigenousbiz_data$cluster)

cluster_map <- leaflet(data = indigenousbiz_data) |>
  addTiles() |>
  setView(lng = -125, lat = 54.5, zoom = 4.5) |>
  addCircleMarkers(
    label = ~business_name,
    radius = 3,
    color = ~pal(cluster),
    stroke = FALSE,
    fillOpacity = 0.7
  ) |>
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~cluster,
    title = "Cluster",
    opacity = 1
  )

cluster_map
```

## Python
```{python analysis_3_py}
#| output: false
db = DBSCAN(eps=1.2, min_samples=8).fit(indigenousbiz_data[["longitude", "latitude"]])

# add cluster labels as new column to the data
indigenousbiz_data["cluster"] = db.labels_.astype(str)

# visualize on map
fig_cluster_map = px.scatter_map(
    indigenousbiz_data,
    lat="latitude",
    lon="longitude",
    color="cluster",
    opacity=0.5,
    zoom=4,
    center={"lat": 54.5, "lon": -128},
    hover_name="business_name",
)

fig_cluster_map = fig_cluster_map.update_layout(
    margin=dict(l=0, r=0, t=30, b=10),
    width=700,
    height=550,
    mapbox_style="open-street-map"
)
```

```{python fig_cluster_map_html}
#| echo: false
html_fig_cluster_map = fig_cluster_map.to_html(include_plotlyjs='cdn', full_html=False)
```

```{r fig_cluster_map_show}
#| echo: false
HTML(py$html_fig_cluster_map)
```

:::

```{r img_save, echo = FALSE, include = FALSE, warning = FALSE, message = FALSE}
library(htmlwidgets)
library(webshot2)

saveWidget(cluster_map, "cluster_map.html", selfcontained = TRUE)
webshot("cluster_map.html", file = "../img/bcindigenousbiz.png", vwidth = 800, vheight = 600)
file.remove("cluster_map.html")
```

#### 5. Treemap of Industry sectors within cluster

::: {.panel-tabset}
## R
```{r analysis_4, warning = FALSE}

## Industry Counts and Treemap Visualization

industry_cluster_counts <- indigenousbiz_data |>
  count(cluster, industry_sector)

ranked_industries <- industry_cluster_counts |>
  group_by(cluster) |>
  mutate(rank = rank(-n, ties.method = "min")) |>
  ungroup()

top_industries <- ranked_industries |>
  mutate(
    industry_sector = if_else(rank <= 3, as.character(industry_sector), "Other")
  ) |>
  group_by(cluster, industry_sector) |>
  summarise(n = sum(n), .groups = "drop") |>
  mutate(
    cluster = factor(cluster),
    industry_sector = factor(industry_sector)
  )

ggplot(top_industries, aes(area = n, fill = industry_sector,
                           label = industry_sector, subgroup = cluster)) +
  geom_treemap() +
  geom_treemap_subgroup_border(color = "white") +
  geom_treemap_text(colour = "white", place = "centre", reflow = TRUE) +
  facet_wrap(~ cluster) +
  coord_cartesian(clip = "off") +
  theme(
    legend.position = "none",
    plot.margin = margin(t = 10, r = 10, b = 10, l = 30)
  ) +
  labs(
    title = "Top Industry Sectors by Geo Cluster",
    fill = "Industry Sector"
  )

```

## Python
```{python analysis_4_py}
industry_cluster_counts = (
    indigenousbiz_data.groupby(by=["cluster", "industry_sector"], observed=True)
    .size()
    .reset_index(name="n")
)

industry_cluster_counts["rank"] = industry_cluster_counts.groupby(
    by="cluster", observed=True
)["n"].rank(method="min", ascending=False)

industry_cluster_counts["industry_sector_label"] = industry_cluster_counts.apply(
    lambda row: row["industry_sector"] if row["rank"] <= 3 else "Other", axis=1
)

top_industries = (
    industry_cluster_counts.groupby(
        by=["cluster", "industry_sector_label"], observed=True
    )
    .agg({"n": "sum"})
    .reset_index()
)

top_industries["hover_text"] = (
    "Cluster: "
    + top_industries["cluster"].astype(str)
    + "<br>Industry: "
    + top_industries["industry_sector_label"].astype(str)
    + "<br>Businesses: "
    + top_industries["n"].astype(str)
)

fig_treemap = px.treemap(
    top_industries,
    path=["cluster", "industry_sector_label"],
    values="n",
    color="industry_sector_label",
    custom_data=["hover_text"],
    title="Top Industry Sectors by Geo Cluster",
)

fig_treemap = fig_treemap.update_traces(
    textinfo="label+value", hovertemplate="%{customdata[0]}<extra></extra>"
)

fig_treemap = fig_treemap.update_layout(margin=dict(l=10, r=10, t=30, b=10))
```

```{python fig_treemap_html}
#| echo: false
html_fig_treemap = fig_treemap.to_html(include_plotlyjs='cdn', full_html=False)
```

```{r fig_treemap_show}
#| echo: false
HTML(py$html_fig_treemap)
```

:::: {.callout-note}
The R and Python cluster labels differ.
::::

:::

### Discussion

This analysis provides a multi faceted view of Indigenous businesses in British Columbia. The statistical and spatial explorations reveal that both regional context and business structure are deeply intertwined with Indigenous entrepreneurship.

The Chi-square test of independence, supported by simulation-based inference due to assumption violations, reveals a statistically significant association between business ownership type and geographic region. This suggests that certain forms of business organization such as community-owned enterprises or private companies may be more prevalent in specific regions, potentially influenced by local governance structures, access to funding, or historical land use patterns.

The spatial clustering using DBSCAN further enriches this narrative by identifying natural groupings of businesses based on their latitude and longitude coordinates. These clusters likely reflect real-world economic and social hubs within the province. The treemap visualization provides a clear breakdown of dominant industry sectors within each spatial cluster, especially after aggregating less-represented sectors into an 'Other' category for interpretability.

In summary, this analysis highlights how Indigenous business patterns in British Columbia are shaped by both geographic and organizational factors. While insightful, the findings should be interpreted with caution given the cross-sectional nature of the data and sensitivity of the clustering method.
