```{r}
#| echo: false
#| warning: false
#| message: false
#| collapse: true
library(reticulate)
if (Sys.getenv("GITHUB_ACTIONS") == "true") {
  # for running in GitHub Actions
  reticulate::use_python(Sys.getenv("RETICULATE_PYTHON"), required = TRUE)
} else {
  # for running locally
  reticulate::use_condaenv("diverse_web_env", required = TRUE)
}

```

### Key Features of the Dataset

Each row of this dataset represents an aggregated demographic group of adults with disabilities and reports their labour force status metrics:

-   **Location** (`geography`) -- The Canadian province or territory of the respondents.
-   **Demographic Information** (`sex`, `age_group`) -- Information on the sex and age bracket of the group.
-   **Disability Profile** (`disability_type`, `global_severity`) -- The specific type of impairment and the overall severity level of the disability (ranging from Mild to Very Severe).
-   **Labour Force Metric** (`labour_force_status`) -- The specific employment metric being reported, such as the employment rate, unemployment rate, or total labour force population count.
-   **Reported Value** (`value`) -- The numerical percentage or population headcount corresponding to the labour force status.

### Purpose and Use Cases

This dataset supports the analysis of:

-   **Employment Equity & Barriers:** Examining how the severity and specific classification of a disability impact an individual's likelihood of securing employment.
-   **Intersectionality in the Labour Market:** Evaluating the combined effects of age, sex, and varying levels of disability severity on workforce participation.
-   **Geographic Disparities:** Comparing employment outcomes for individuals with disabilities across different Canadian provinces and territories to identify regional inequalities.
-   **Predictive Modeling:** Building fractional regression baselines to estimate expected employment probabilities based on a combination of medical and demographic factors.
-   **Institutional & Policy Insights:** Informing the distribution of federal and provincial social assistance (such as the Canada Disability Benefit), adjusting earnings exemption limits, and designing targeted vocational rehabilitation programs.

## Case Study

### Objective

Equitable employment is a cornerstone of financial independence and social inclusion for adults living with disabilities. In Canada, federal and provincial governments provide essential financial assistance, such as the [Canada Disability Benefit (CDB)](https://www.canada.ca/en/services/benefits/disability/canada-disability-benefit.html) and provincial support programs, to help bridge the income gap. However, the reliance on these support systems and the ability to find sustainable employment are heavily influenced by a complex intersection of demographic factors.

To better understand the structural barriers within the Canadian labour market, this case study leverages Statistics Canada data to evaluate employment outcomes. We focus on several key questions:

-   How does the global severity of a disability impact the likelihood of employment?
-   Are there significant disparities in employment rates across different age groups and sexes?
-   To what extent does geographical location (province or territory) affect labour force inclusion?

By conducting exploratory data analysis and fitting a regression model, we aim to statistically quantify the primary drivers of employment for individuals with disabilities. This analysis provides an educational framework for understanding how data can inform public policy, optimize vocational rehabilitation resources, and highlight regional or demographic gaps in the current social.

### Analysis

**Loading Libraries**

::: panel-tabset
## R

```{r load-libraries_r, results = "hide", warning = FALSE, message = FALSE}
# Data
library(diversedata)      # Diverse Data Hub datasets

# Core libraries
library(tidyverse)
library(knitr)

# Statistical Modeling
library(broom)

```

## Python

```{python load-libraries_py, results = "hide", warning = FALSE, message = FALSE}
import pandas as pd
import numpy as np
from IPython.display import Markdown
import statsmodels.api as sm
import statsmodels.formula.api as smf
import matplotlib.pyplot as plt
import seaborn as sns

```
:::

#### 1. Data Cleaning & Processing

**Note:** When working with aggregated government data, metrics like "Employment Rate" are often pre-calculated percentages. To use these in a logistical or fractional regression model, we must convert them into a 0-1 proportion.

::: panel-tabset
## R

```{r data_prep_r}
# Reading data
labour_force <- labourforcewithdisabilities

# Filter for the target disability category to avoid double-counting
df_all_disabilities <- labour_force |>
  filter(disability_type == 'All disability types')

# Separate the employment rates from the population counts
df_rates <- df_all_disabilities |>
  filter(labour_force_status == 'Employment Rate') |>
  rename(employment_rate = value)

df_pop <- df_all_disabilities |>
  filter(labour_force_status == 'Total labour force') |>
  rename(population_weight = value)

# Merge to align rates with their respective population weights
merge_cols <- c('geography', 'sex', 'age_group', 'global_severity')

labour_force_clean <- df_rates |>
  select(all_of(merge_cols), employment_rate) |>
  inner_join(df_pop |> select(all_of(merge_cols), population_weight), by = merge_cols) |>
  drop_na(employment_rate, population_weight) |>
  # Convert rate to a 0-1 proportion for the Fractional Logit model
  mutate(employment_proportion = employment_rate / 100.0)

# Preview data
labour_force_clean |>
  head() |>
  kable()

```

## Python

```{python data_prep_py}
# Reading data
# Will need to switch data loading step when PyPI is updated!
#labour_force = dd.load_data("labourforcewithdisabilities")
labour_force = pd.read_csv("../../data/clean/labourforcewithdisabilities.csv")

# Filter for the target disability category to avoid double-counting
df_all_disabilities = labour_force[labour_force['disability_type'] == 'All disability types']

# Separate the employment rates from the population counts
df_rates = df_all_disabilities[df_all_disabilities['labour_force_status'] == 'Employment Rate'].copy()
df_pop = df_all_disabilities[df_all_disabilities['labour_force_status'] == 'Total labour force'].copy()

# Rename the 'value' columns before merging
df_rates = df_rates.rename(columns={'value': 'employment_rate'})
df_pop = df_pop.rename(columns={'value': 'population_weight'})

# Merge to align rates with their respective population weights
merge_cols = ['geography', 'sex', 'age_group', 'global_severity']
labour_force_clean = pd.merge(df_rates[merge_cols + ['employment_rate']], 
                    df_pop[merge_cols + ['population_weight']], 
                    on=merge_cols, 
                    how='inner')

# Drop any rows with suppressed or missing data
labour_force_clean = labour_force_clean.dropna(subset=['employment_rate', 'population_weight'])

# Convert rate to a 0-1 proportion for the Fractional Logit model
labour_force_clean['employment_proportion'] = labour_force_clean['employment_rate'] / 100.0

Markdown(labour_force_clean.head().to_markdown())


```
:::

#### 2. Exploratory Data Analysis

Before applying predictive modeling or inferential tests, visual exploration provides an intuitive baseline of how employment rates vary across individual demographic factors.

------------------------------------------------------------------------

**Employment Rate by Severity Level**

This visualization evaluates the relationship between disability severity and employment outcomes. It aggregates reported employment rates across four global severity tiers: Mild, Moderate, Severe, and Very Severe.

::: panel-tabset
## R

```{r eda_severity_r, fig.width=8, fig.height=4.5}
sev_order <- c("Mild", "Moderate", "Severe", "Very Severe")

ggplot(labour_force_clean, aes(
  x = employment_rate, 
  y = factor(global_severity, levels = rev(sev_order)), # rev() plots Mild at the top
  fill = factor(global_severity, levels = sev_order)    # Maps light-to-dark logically
)) +
  stat_summary(fun = "mean", geom = "bar", width = 0.6) +
  scale_fill_brewer(palette = "Reds") +
  labs(
    title = "Mean Employment Rate by Severity Level",
    x = "Employment Rate (%)",
    y = "Global Severity"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank()
  )

```

## Python

```{python eda_severity_py}
sev_order = ["Mild", "Moderate", "Severe", "Very Severe"]

plt.figure(figsize=(8, 4.5))
sns.barplot(
    data=labour_force_clean,
    x="employment_rate",
    y="global_severity",
    hue="global_severity",
    order=sev_order,
    hue_order=sev_order,
    errorbar=None,
    palette="Reds",
    legend=False,
    width=0.6,
)

plt.title("Mean Employment Rate by Severity Level", fontsize=13, fontweight="bold", pad=12)
plt.xlabel("Employment Rate (%)", fontsize=11)
plt.ylabel("Global Severity", fontsize=11)
plt.grid(True, alpha=0.5)
plt.tight_layout()
plt.show()

```
:::

A clear negative trajectory emerges between impairment severity and labour market participation. Individuals classified with Mild disabilities experience employment rates approaching 70%, whereas the likelihood of holding a job declines sharply as severity transitions to Moderate, Severe, and Very Severe.

------------------------------------------------------------------------

**Employment Rate by Age Group**

This plot examines labour force engagement across different stages of working life, from youth entering the workforce to older adults approaching retirement.

::: panel-tabset
## R

```{r eda_age_r, fig.width=8, fig.height=4.5}
age_order <- c("15 to 24 years", "25 to 34 years", "35 to 44 years", "45 to 54 years", "55 to 64 years")

ggplot(labour_force_clean, aes(
  x = employment_rate, 
  y = factor(age_group, levels = rev(age_order)), 
  fill = factor(age_group, levels = age_order)
)) +
  stat_summary(fun = "mean", geom = "bar", width = 0.6) +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Mean Employment Rate by Age Group",
    x = "Employment Rate (%)",
    y = "Age Group"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank()
  )

```

## Python

```{python eda_age_py}
age_order = [
    "15 to 24 years", "25 to 34 years", 
    "35 to 44 years", "45 to 54 years", "55 to 64 years"
]

plt.figure(figsize=(8, 4.5))
sns.barplot(
    data=labour_force_clean,
    x="employment_rate",
    y="age_group",
    hue="age_group",
    order=age_order,
    hue_order=age_order,
    errorbar=None,
    palette="Blues",
    legend=False,
    width=0.6,
)

plt.title("Mean Employment Rate by Age Group", fontsize=13, fontweight="bold", pad=12)
plt.xlabel("Employment Rate (%)", fontsize=11)
plt.ylabel("Age Group", fontsize=11)
plt.grid(True, alpha=0.5)
plt.tight_layout()
plt.show()

```
:::

Employment rates follow a distinct bell curve across age categories. Working-age adults in the core brackets (25 to 54 years) maintain the highest average employment rates. Conversely, youth (15 to 24 years) and older adults (55 to 64 years) experience substantial penalties, reflecting transition barriers for new labour market entrants and early exits from the workforce near retirement age.

------------------------------------------------------------------------

**Employment Rate by Sex**

This comparison breaks down average employment rates between males and females in the dataset to highlight gender-based participation gaps.

::: panel-tabset
## R

```{r eda_sex_r, fig.width=8, fig.height=3.5}
sex_order <- c("Females", "Males")

ggplot(labour_force_clean, aes(
  x = employment_rate, 
  y = factor(sex, levels = rev(sex_order)), 
  fill = factor(sex, levels = sex_order)
)) +
  stat_summary(fun = "mean", geom = "bar", width = 0.5) +
  scale_fill_brewer(palette = "Greens") +
  labs(
    title = "Mean Employment Rate by Sex",
    x = "Employment Rate (%)",
    y = "Sex"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank()
  )

```

## Python

```{python eda_sex_py}
sex_order = ["Females", "Males"]

plt.figure(figsize=(8, 3.5))
sns.barplot(
    data=labour_force_clean,
    x="employment_rate",
    y="sex",
    hue="sex",
    order=sex_order,
    hue_order=sex_order,
    errorbar=None,
    palette="Greens",
    legend=False,
    width=0.5,
)

plt.title("Mean Employment Rate by Sex", fontsize=13, fontweight="bold", pad=12)
plt.xlabel("Employment Rate (%)", fontsize=11)
plt.ylabel("Sex", fontsize=11)
plt.grid(True, alpha=0.5)
plt.tight_layout()
plt.show()

```
:::

Males display a marginally higher mean employment rate compared to females. While this gap is narrower than the disparities observed across age brackets or severity levels, it indicates that gender intersects with disability to shape workforce attachment.

------------------------------------------------------------------------

**Employment Rate by Province and Territory**

This chart displays geographic variation in employment rates across Canadian provinces and territories, ordered from highest to lowest mean employment rate.

::: panel-tabset
## R

```{r eda_geo_r, fig.width=8, fig.height=6}
ggplot(labour_force_clean, aes(
  x = employment_rate, 
  y = reorder(geography, employment_rate, FUN = mean), 
  fill = reorder(geography, employment_rate, FUN = mean)
)) +
  stat_summary(fun = "mean", geom = "bar", width = 0.6) +
  scale_fill_viridis_d(option = "magma") +
  labs(
    title = "Mean Employment Rate by Province/Territory",
    subtitle = "Ordered by Mean Employment Rate",
    x = "Employment Rate (%)",
    y = "Province / Territory"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "gray40", size = 10),
    panel.grid.minor = element_blank()
  )

```

## Python

```{python eda_geo_py}
# Calculate means and order ascending to match R's color scaling
geo_means = labour_force_clean.groupby("geography")["employment_rate"].mean().sort_values(ascending=True)
geo_order_asc = geo_means.index.tolist()

# Reverse for the y-axis so the highest value sits at the top of the plot
geo_order_desc = geo_order_asc[::-1]

plt.figure(figsize=(8, 6))
sns.barplot(
    data=labour_force_clean,
    x="employment_rate",
    y="geography",
    hue="geography",
    order=geo_order_desc,       # Plot highest at top
    hue_order=geo_order_asc,    # Map magma lowest-to-highest to match R
    errorbar=None,
    palette="magma",
    legend=False,
    width=0.6,
)

plt.title("Mean Employment Rate by Province/Territory", fontsize=13, fontweight="bold", pad=15, loc="left")
plt.xlabel("Employment Rate (%)", fontsize=11)
plt.ylabel("Province / Territory", fontsize=11)
plt.grid(True, alpha=0.5)
plt.tight_layout()
plt.show()

```
:::

Regional variations highlight differences in economic conditions and social support frameworks. Western provinces such as Saskatchewan and Alberta report higher average employment rates, while Quebec and Atlantic provinces tend to show lower baseline figures. Although northern territories (Yukon and Nunavut) appear at the top of the raw rankings, these estimates stem from smaller sample sizes with high suppression rates and require careful interpretation through weighted modeling.


#### 3. Weighted Fractional Logit Regression

**Note:** We use a Generalized Linear Model (GLM) with a `Binomial` (or `quasibinomial`) family to handle fractional response variables (proportions between 0 and 1). By passing our population headcounts into the `weights` argument, the model evaluates the dataset as millions of surveyed individuals rather than treating each aggregated row equally. 

**Setting Reference Categories**

Before running a regression, the model needs a baseline to compare everything against. By explicitly setting reference groups, our final coefficients will clearly show the effect relative to: `Ontario`, `Males`, `25 to 34 years`, and `Mild` severity.

::: panel-tabset
## R

```{r set_references_r}
# Explicitly setting reference groups for the categorical variables directly in the dataframe
labour_force_clean <- labour_force_clean |>
  mutate(
    geography = relevel(factor(geography), ref = "Ontario"),
    sex = relevel(factor(sex), ref = "Males"),
    age_group = relevel(factor(age_group), ref = "25 to 34 years"),
    global_severity = relevel(factor(global_severity), ref = "Mild")
  )

```

## Python

```{python set_references_py}
# Define formula explicitly setting the reference group for each demographic
formula = (
    "employment_proportion ~ "
    "C(geography, Treatment(reference='Ontario')) + "
    "C(sex, Treatment(reference='Males')) + "
    "C(age_group, Treatment(reference='25 to 34 years')) + "
    "C(global_severity, Treatment(reference='Mild'))"
)

```

:::

**Fitting the Model**

Next, we feed our prepared variables into the GLM. Because the outcome is a proportion, a standard linear model would be inappropriate; instead, the logit link transforms the data into log-odds.

::: panel-tabset

## R

```{r fit_glm_r}
# Fit the GLM with quasibinomial family (to handle overdispersion) and population weights
weighted_glm <- glm(
  employment_proportion ~ geography + sex + age_group + global_severity,
  data = labour_force_clean,
  family = quasibinomial(link = "logit"),
  weights = population_weight
)

```

## Python

```{python fit_glm_py}
# Fit a GLM with Binomial family, population weights, 
# and Pearson's chi-squared scale to account for overdispersion (Quasibinomial equivalent)
weighted_glm_py = smf.glm(
    formula=formula, 
    data=labour_force_clean, 
    family=sm.families.Binomial(), 
    var_weights=labour_force_clean['population_weight']
).fit(scale="X2")

```

:::

**Formatting the Output**

Raw model output can be dense and difficult to read. In this final step, we clean up the summary statistics so they are presentation-ready.

::: panel-tabset

## R

```{r display_results_r}
# Tidy the model output, round numbers for readability, and render a clean table
tidy(weighted_glm) |>
  mutate(across(where(is.numeric), ~round(.x, 3))) |>
  kable()

```

## Python

```{python display_results_py}
# Print the formatted statsmodels summary table
print(weighted_glm_py.summary())

```

:::


### Discussion

The model establishes a clear reference point: a Male, aged 25 to 34 years, located in Ontario, with a Mild disability. The intercept log-odds ($0.9377$) converts to a baseline expected employment rate of roughly **71.8%**.

Holding all other variables constant, we can isolate specific systemic barriers:

* **Severity Penalty:** Moving from a Mild disability to Moderate ($-0.4496$), Severe ($-0.9263$), or Very Severe ($-1.5372$) strictly and significantly decreases the likelihood of employment.
* **Age Penalty:** Youth aged 15 to 24 face the steepest overall age penalty in the model ($-1.2410$). Older adults aged 55 to 64 face a substantial drop ($-0.6809$) as they approach retirement.
* **Data Sparsity:** Nunavut returns a high $p$-value ($0.947$), validating our EDA assumption that the territory's sample size is too volatile to draw reliable inferential conclusions without wider confidence intervals.

Understanding these exact log-odds penalties allows educational and governmental institutions to see exactly where current social safety nets and vocational programs are succeeding, and where they are falling short.