```{r}
#| echo: false
#| warning: false
#| message: false
#| collapse: true
library(reticulate)
if (Sys.getenv("GITHUB_ACTIONS") == "true") {
  # for running in GitHub Actions (otherwise reticulate can't find diverse_web_env)
  reticulate::use_python(Sys.getenv("RETICULATE_PYTHON"), required = TRUE)
} else {
  # for running locally
  reticulate::use_condaenv("diverse_web_env", required = TRUE)
}
```

### Key Features of the Dataset

Each row of this dataset represents a group of graduates and reports the group's median income 2 years after graduation, along with demographic information about the group of graduates:

- **Year** (`year`) -- The year in which income was reported
- **Location** (`location`) -- The geographical location of the graduates
- **Educational Information** (`educational_qualification`, `field_of_study`) -- Information on the type and focus of the degree received by the group of graduates
- **Demographic Information** (`gender`, `age_group`, `population_group`) -- Demographic information on the gender, age, and population group (similar to ethnicity) of the group of graduates
- **Median Income** (`median_income`) -- The median income of the group of graduates
- **Number of Graduates** (`number_graduates`) -- The number of individuals that make up the group

### Purpose and Use Cases
This dataset supports the analysis of: 

- **Demographic Equity & Disparities:** Examining wage gaps and income distribution across gender and population groups.
- **Educational Return on Investment (ROI):** Evaluating how different educational qualifications and credential levels impact post-graduation incomes.
- **Impact of Field of Study:** Comparing incomes across various academic disciplines, from professional and technical trades to the humanities and sciences.
- **Predictive Modeling:** Building regression or machine learning baselines to estimate expected post-graduate income based on a combination of educational and demographic factors.
- **Institutional & Policy Insights:** Informing career advisory strategies, educational planning, and equity initiatives for prospective students.

## Case Study

### Objective
Equitable incomes for recent graduates are essential for inclusive labor markets and educational planning. This dataset provides information on post-graduation median incomes in Canada, allowing us to examine how factors like gender, ehtnicity, education type, and field of study intersect and impact incomes. 

To explore these dynamics, we focus on several key questions: 

- Is there evidence of income disparity by gender or ethnicity in recent graduates in Canada?
- How do educational credentials and fields of study shape incomes across different groups? 
- To what extent can we predict post-graduation income using these academic and demographic factors? 

Through statistical testing and predictive modeling, we aim to uncover structural wage gaps and identify the primary drivers of post-grad earnings.

### Analysis

**Loading Libraries**

::: {.panel-tabset}
## R
```{r load-libraries, results = "hide", warning = FALSE, message = FALSE}
# Data
library(diversedata)      # Diverse Data Hub datasets

# Core libraries
library(tidyverse)        
library(lubridate)      
library(stringr)

# Hypothesis testing
library(car)
library(FSA)

# Tables & reporting
library(gt)               
library(kableExtra)       
       
```

## Python
```{python load-libraries_py, results = "hide", warning = FALSE, message = FALSE}
#import diversedata as dd
#import numpy as np
#import pandas as pd
```
:::

#### 1. Data Cleaning & Processing
To clean the data, first we use regex to clean the values in `field_of_study` and remove unnecessary numerical markers, like `[1]`. Then, we convert relevant columns to factor type. 

::: {.panel-tabset}
## R
```{r data_load}
income_data <- postgradincome |>
  mutate(
    field_of_study = str_remove(field_of_study, "\\s*\\[\\d+\\]$"),
    across(
      c(location, educational_qualification, field_of_study, gender, age_group, population_group), 
      as.factor
    )
  ) 

# preview data
income_data |>
  head() |>
  kable()

```

:::

#### 2. Exploratory Data Analysis

::: {.panel-tabset}
## R
**Median Income by Gender**
```{r eda_1_r}

library(ggdist)

ggplot(income_data, aes(
  y = gender, 
  x = median_income)) +
  stat_halfeye(aes(weight = number_graduates), 
               adjust = 1.5, 
               justification = -0.1, 
               .width = 0, 
               point_colour = NA, 
               alpha = 0.6,
               fill = "darkgreen") +
  geom_boxplot(aes(weight = number_graduates), 
               width = 0.15, 
               outlier.shape = NA, 
               alpha = 0.7, 
               color = "black") +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Median Income Distribution by Gender",
    y = "Gender",
    x = "Median Income ($)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank())

```
**Median Income by Population Group**

```{r eda_2_r}
#| fig-width: 8
#| fig-height: 6

ggplot(income_data, aes(
  y = fct_reorder(population_group, median_income, .fun = median, .na_rm = TRUE), 
  x = median_income)) +
  stat_halfeye(aes(weight = number_graduates), 
               adjust = 1.5, 
               justification = -0.1, 
               .width = 0, 
               point_colour = NA, 
               alpha = 0.6,
               fill = "darkgreen") +
  geom_boxplot(aes(weight = number_graduates), 
               width = 0.15, 
               outlier.shape = NA, 
               alpha = 0.7, 
               color = "black") +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Median Income Distribution by Population Group",
    subtitle = "Ordered by Median Income",
    y = "Population Group",
    x = "Median Income ($)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank())

```

**Median Income by Level of Study**
```{r eda_3_r}
#| fig-width: 8
#| fig-height: 6

ggplot(income_data, aes(
  y = fct_reorder(educational_qualification, median_income, .fun = median, .na_rm = TRUE), 
  x = median_income)) +
  stat_halfeye(aes(weight = number_graduates), 
               adjust = 1.5, 
               justification = -0.1, 
               .width = 0, 
               point_colour = NA, 
               alpha = 0.6,
               fill = "darkgreen") +
  geom_boxplot(aes(weight = number_graduates), 
               width = 0.15, 
               outlier.shape = NA, 
               alpha = 0.7, 
               color = "black") +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_discrete(labels = scales::label_wrap(width = 35)) +
  labs(
    title = "Median Income Distribution by Educational Qualification",
    subtitle = "Ordered by Median Income",
    y = "Educational Qualification",
    x = "Median Income ($)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank())

```

**Median Income by Field of Study**
```{r eda_4_r}
#| fig-width: 8
#| fig-height: 6

ggplot(income_data, aes(
  y = fct_reorder(field_of_study, median_income, .fun = median, .na_rm = TRUE), 
  x = median_income)) +
  stat_halfeye(aes(weight = number_graduates), 
               adjust = 1.5, 
               justification = -0.1, 
               .width = 0, 
               point_colour = NA, 
               alpha = 0.6,
               fill = "darkgreen") +
  geom_boxplot(aes(weight = number_graduates), 
               width = 0.15, 
               outlier.shape = NA, 
               alpha = 0.7, 
               color = "black") +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_discrete(labels = scales::label_wrap(width = 35)) +
  labs(
    title = "Median Income Distribution by Field of Study",
    subtitle = "Ordered by Median Income",
    y = "Field of Study",
    x = "Median Income ($)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank())

```
:::

#### 3. Hypothesis Testing

We want to test if different groups of graduates have significantly different median incomes. We will investigate both gender and population group as a proxy for ethnicity, but let's begin with gender. 

**Hypothesis Testing: Gender**

Our hypotheses are: 

* **Null Hypothesis ($H_0$):** The distributions of median income are identical for both genders.
* **Alternative Hypothesis ($H_a$):** The distributions of median income are not identical for both genders.

Before we can perform our ANOVA hypothesis test, we need to check if the required assumptions are met: 

1. Independence of Observations

As the dataset includes one row per discrete group of graduates, and there are no repetitions, this dataset satisfactorily fulfills the required independence of observations. 

2. Normality of Residuals
We build a Q-Q plot, where, if the residuals are normally distributed, we would expect the black points to lie close to the red diagonal line. 

::: {.panel-tabset}
## R
```{r qq_plot_r}

anova_gender_model <- aov(median_income ~ gender, data = income_data)
qqnorm(residuals(anova_gender_model))
qqline(residuals(anova_gender_model), col = "red", lwd = 2)

```
:::

Above, we can see that the residuals are not normally distributed, as the points deviate from the red line on both extremes. 

We can additionally test this normality by making a density plot of the residuals, and visually inspecting the shape of the curve. 

::: {.panel-tabset}
## R
```{r residuals_r}

ggplot(data.frame(resids = residuals(anova_gender_model)), aes(x = resids)) +
  geom_density(fill = "#7EA8F8", alpha = 0.6) +
  labs(
    title = "Density Plot of ANOVA Residuals",
    x = "Residuals",
    y = "Density"
  ) +
  theme_minimal()

```
:::
We can see that the residuals are not normally distributed above, but instead are right-skewed. This is typical in income datasets, as most people have a low to moderate income, while a small number of people earn very high amounts, as discussed in [Passion Driven Statistics](https://alanarnholt.github.io/PDS-Bookdown2/skewed-right-distributions.html). 

3. Homogeneity of Variances - Levene's test

For ANOVA, we also need to check if the different groups have comparable amounts of variance. 

::: {.panel-tabset}
## R
```{r variance_r}

leveneTest(median_income ~ gender, data = income_data)

```
:::

We can see that the variances are not homogeneous in our datset, as the p-value from the Levene's test is smaller than 0.05. 

Since the residuals are not normally distributed and the variances are not homogeneous, we cannot use an ANOVA test. Instead, we will use a non-parametric Kruskal-Wallis test. 

::: {.panel-tabset}
## R
```{r kruskal_r}

kruskal.test(median_income ~ gender, data = income_data)

```
:::
Since p-value is smaller than 0.05, we can reject the null hypothesis in favor of the alternative, and conclude that there is statistical evidence for a difference in median income between men and women in our dataset. 

**Hypothesis Testing: Population Group**

We can do the same process for income by population group. Our hypotheses are:   

* **Null Hypothesis ($H_0$):** The distributions of median income are identical across all population groups.
* **Alternative Hypothesis ($H_a$):** The distributions of median income are not identical across all population groups.


First we need to re-check our assumptions. 

Assumption checks:

1. Independence of Observations

As the dataset includes one row per discrete group of graduates, and there are no repetitions, this dataset satisfactorily fulfills the required independence of observations. 

2. Normality of Residuals

::: {.panel-tabset}
## R
```{r qq_plot_popgroup_r}

anova_popgroup_model <- aov(median_income ~ population_group, data = income_data)
qqnorm(residuals(anova_popgroup_model))
qqline(residuals(anova_popgroup_model), col = "red", lwd = 2)


```
:::

::: {.panel-tabset}
## R
```{r residuals_popgroup_r}

ggplot(data.frame(resids = residuals(anova_popgroup_model)), aes(x = resids)) +
  geom_density(fill = "#7EA8F8", alpha = 0.6) +
  labs(
    title = "Density Plot of ANOVA Residuals",
    x = "Residuals",
    y = "Density"
  ) +
  theme_minimal()

```
:::

Again, from both the Q-Q plot and the residuals density plot, we can see that the residuals are not normally distributed. 

3. Homogeneity of Variances - Levene's test

::: {.panel-tabset}
## R
```{r variance_popgroup_r}

leveneTest(median_income ~ population_group, data = income_data)

```
:::

Similarly, since the p-value is less than 0.05, we can conclude that the variances are heterogeneous, and this assumption is violated. 

Therefore, since the residuals are not normally distributed and the variances are not homogeneous, we cannot use ANOVA and instead will use a Kruskal-Wallis test. 

::: {.panel-tabset}
## R
```{r kruskal_popgroup_r}

kruskal.test(median_income ~ population_group, data = income_data)

```
:::

Since p-value is smaller than 0.05, we can reject the null hypothesis in favor of the alternative, that the distributions of median income are not identical across all population groups. To find out which groups differ, we need to conduct a Dunn's Post-Hoc test with a Bonferroni correction for multiple comparisons. 

::: {.panel-tabset}
## R
```{r dunns_popgroup_r}

# Run Dunn's test with Bonferroni correction
dunnTest(median_income ~ population_group, data = income_data, method = "bonferroni")

```
:::

From both the Kruskal-Wallis and the Dunn's Post-Hoc test, we can see that there are statistically significant differences in median income distributions between different population groups. Notable differences appear around comparing "Not a visible minority" to many groups, including "Arab", "Black", "Latin American", etc. 


#### 4. Predictive Model

To understand the factors driving median income and identify variations across groups, we will build a linear regression model. Because our dataset is aggregated, where each row represents a group of individuals rather than a single person, we need to use the number of individuals contained in the `number_graduates` column as the weight of each row in data splitting as well as model training and evaluation. 

Our modeling workflow is structured as follows:

* **Data Splitting:** We partition the weighted dataset into training and test sets to ensure unbiased evaluation.
* **Coefficient & Group Analysis:** We examine the model's learned coefficients and their statistical significance to uncover key drivers and differences between groups.
* **Performance Evaluation:** We assess model generalizability by calculating the root mean squared error (RMSE) of our predictions on the test set.

::: {.panel-tabset}
## R
```{r split_data_weighted_r}
library(rsample)
set.seed(42)

# Split dataset to train and test sets using stratified splitting
# Ensures small and large weight rows are evenly distributed between train and test
data_split <- initial_split(income_data, prop = 0.8, strata = number_graduates)

# Create the training and testing sets
income_train_weighted <- training(data_split)
income_test_weighted  <- testing(data_split)

```
:::

::: {.panel-tabset}
## R
```{r weighted_model_r}
# Fit a weighted linear model
model_lm_weighted <- lm(median_income ~ educational_qualification + population_group + gender + field_of_study, 
               data = income_train_weighted,
               weights = number_graduates)

# View coefficients and summary
summary(model_lm_weighted)
```
:::

Overall, this model performed reasonably well. The Adjusted R-squared ($0.6491$) indicates that this model can explain 64.91% of the variance in median post-graduation income. This number, while not extremely high, provides a good baseline prediction for a complex social and economic dataset. The F-statistic also indicates that this model is doing well. With a p-value of $p < 2.2 \times 10^{-16}$), the model as a whole is statistically significant compared to an intercept-only model, which indicates that the explanatory variables used here do a decent job of explaining income differences. 

We can interpret the coefficients from our model as well. The intercept ($\$31,044$) represents the baseline predicted median income for the reference group (i.e. graduates in the group of "Man" and "Arab", with education in "Career, technical or professional training certificate", and "Agriculture, natural resources and conservation"). All coefficients represent the expected difference in median income relative to the reference category, holding all other variables constant:

* **Gender:** Being a woman, compared to being a man, is associated with a median income that is $\$4,884$ lower, while holding all other variables constant. 
* **Population Group:** Several groups show significant differences relative to the baseline group, "Arab". For example, "Not a visible minority" shows an estimated increase of $\$7,136$, while groups like Japanese, Multiple visible minorities, and West Asian do not show statistically significant differences ($p > 0.05$) after controlling for education, gender, and field of study.
* **Educational qualification:** Higher degrees show higher median incomes, as is expected. For example, a Doctoral degree is associated with an increase of $\$54,592$, and a Professional degree with an increase of about $\$56,240$, while holding all other variables constant. 
* **Field of Study:** Fields like Health and related fields ($\approx +\$14,448$) and Architecture and engineering ($\approx +\$12,102$) show large positive coefficients, whereas fields like Visual and performing arts ($\approx -\$17,637$) and Humanities ($\approx -\$16,249$) show substantial negative adjustments compared to the baseline field of "Agriculture, natural resources and conservation". 

::: {.panel-tabset}
## R
```{r weighted_model_eval_r}
# Predict on test data
test_results_weighted <- income_test_weighted %>%
  mutate(.pred = predict(model_lm_weighted, newdata = income_test_weighted))

# Calculate weighted RMSE
test_results_weighted %>%
  summarize(
    weighted_mse = sum(number_graduates * (median_income - .pred)^2, na.rm = TRUE) / sum(number_graduates, na.rm = TRUE),
    weighted_rmse = sqrt(weighted_mse)
  )

# Display income range for comparison
cat(sprintf("Minimum median income: $%s\n", format(min(income_data$median_income), big.mark = ",")))
cat(sprintf("Maximum median income: $%s\n", format(max(income_data$median_income), big.mark = ",")))
cat(sprintf("Mean median income: $%s\n", format(round(mean(income_data$median_income), 0), big.mark = ",")))

```
:::

To further evaluate our model's generalizability, we can generate predictions on the test set and calculate the weighted root mean squared error (RMSE). The resulting test RMSE is $\$12,135$. 

To put this error into perspective, we can compare it against the broader distribution of median incomes in our dataset, which spans from $\$3,900$ to $\$270,200$ with a mean of $\$62,173$. An RMSE of roughly $\$12,000$. represents about 20% of the average income value. This indicates that while there is room for higher precision, the model captures the overall trend effectively without large-scale deviation.

To ground our evaluation in a concrete example, we can inspect a single row from our test set. By comparing the model's predicted income against the true reported value for this group, we can see how the model performs on an individual data point while accounting for its corresponding sample weight.

::: {.panel-tabset}
## R
```{r single_example_r}

# Select a single row from the test set (e.g. the first row)
example_row <- income_test_weighted |> 
  slice(2)

# Generate the prediction for this specific row
example_pred <- predict(model_lm_weighted, newdata = example_row)

# Calculate difference in predicted and true median incomes
ex_true <- example_row$median_income
ex_diff <- abs(ex_true - example_pred)

# Display the attributes and values of our test case
cat(sprintf("Educational Qualification: %s\n", example_row$educational_qualification))
cat(sprintf("Field of Study: %s\n", example_row$field_of_study))
cat(sprintf("Gender: %s\n", example_row$gender))
cat(sprintf("Population Group: %s\n", example_row$population_group))
cat(sprintf("Number of Graduates in Group (Weight): %s\n", format(example_row$number_graduates, big.mark = ",")))
cat(sprintf("True Median Income: $%s\n", format(round(ex_true, 2), big.mark = ",")))
cat(sprintf("Predicted Median Income: $%s\n", format(round(example_pred, 2), big.mark = ",")))
cat(sprintf("Absolute Error: $%s\n", format(round(ex_diff, 2), big.mark = ",")))
```
:::

Looking at this specific test case of a group of graduates from our test set, the true median income and our model's predicted median income give us a clear concrete example of how the model performs on an individual level. 

Given that our test set RMSE is $12,135, individual discrepancies like this align fairly well with our overall error metrics. It illustrates that while the model captures the structural baseline and general trajectory of graduate earnings quite well, localized variance among specific demographic and educational intersections can still produce deviations of around five to twenty thousand dollars.




#### Interaction Effects

Above, we used a solely additive model, meaning that we assumed that the effects of each explanatory variable on the income prediction were independent of each other. Let's investigate an interactive model, which allows the effect of one variable to depend on the level of another. We will examine whether gender wage disparities vary across different academic areas, by incorporating interaction terms between gender and field of study into our model. This allows us to test whether specific fields exhibit wider or narrower wage gaps between men and women, after controlling for educational qualifications and population groups. It is important to note here, that we are restricting this investigation to only interaction terms between two variables, gender and field of study, as including interaction terms between all four explanatory variables can lead to severe overfitting, singularities due to uncommon combinations, and uninterpretable parameter explosion. These issues, and methods to work around them, are discussed further [here](https://medium.com/@TingyuZou/when-analyzing-complex-data-researchers-often-encounter-relationships-that-cannot-be-captured-by-1d9dccf14583). 


::: {.panel-tabset}
## R
```{r inter_model_r}
# Fit a weighted linear model with interaction effects
model_lm_inter <- lm(median_income ~ (educational_qualification + population_group + gender * field_of_study), 
               data = income_train_weighted,
               weights = number_graduates)

# View coefficients and summary
summary(model_lm_inter)
```
:::

::: {.panel-tabset}
## R
```{r inter_model_eval_r}
# Predict on test data
test_results_inter <- income_test_weighted %>%
  mutate(.pred = predict(model_lm_inter, newdata = income_test_weighted))

# Calculate weighted RMSE
test_results_inter %>%
  summarize(
    weighted_mse = sum(number_graduates * (median_income - .pred)^2, na.rm = TRUE) / sum(number_graduates, na.rm = TRUE),
    weighted_rmse = sqrt(weighted_mse)
  )

# Display income range for comparison
cat(sprintf("Minimum median income: $%s\n", format(min(income_data$median_income), big.mark = ",")))
cat(sprintf("Maximum median income: $%s\n", format(max(income_data$median_income), big.mark = ",")))
cat(sprintf("Mean median income: $%s\n", format(round(mean(income_data$median_income), 0), big.mark = ",")))

```
:::

The inclusion of the gender - field of study interactions modestly improves model fit, raising the adjusted R-squared to $0.6594$. While a general baseline gender penalty persists across many fields, where women earn an estimated $\$7,205$ less than men holding other factors constant, the magnitude of this gap varies significantly by discipline:

* **Exacerbated Disparities in Tech & Trades:** In fields like **mathematics, computer, and information sciences** ($\-$ $\$5,435$ interaction) and **personal, protective, and transportation services** ($\-$ $\$9,159$ interaction), the negative interaction coefficients indicate a structurally wider gender gap.
* **Neutralization in Health:** Conversely, **health and related fields** exhibit a significant positive interaction ($\approx +\$10,597$, $p < 0.001$), which effectively neutralizes or reverses the baseline gender penalty in those specific credentials. 

These findings suggest that post-graduation wage disparities are not uniform across the labor market, but are instead heavily mediated by the specific career paths graduates pursue. With further analysis and investigation, other interaction terms may be significant, and provide more detailed insights into wage disparities. 


### Discussion
Results from both the hypothesis testing and model training suggest that income disparity continues to prevail by both gender and ethnicity in the current day. 




